---
  title: "Flow Control"
output: html_notebook
---


```{r}
X <- 21
Xvec <- c(1:10, "string")
```

```{r}
if(X > 100){ #Is X greater than 100?
  print("X > 100") #If True
} else {
  print("X <= 100") #If False
}
```

```{r}
if(X > 0){print ("yup")}
```

```{r}
ifelse(X > 100, "X > 100", "X <= 100")
```

```{r}
# Nested if and else statements account for more outcomes (like a branching tree for every __ statement)

if(X > 100){
  print("X > 100")
  if(X > 200){
    print("X > 200")
  }
} else {
  if(X == 100){
    print("X = 100")
  } else {
    print("X < 100")
  }
}

```

## Loops do the same thing repeatedly until the condition is met. They use the basic format for(){}


```{r}
for (i in 1:5){
  print(paste(X,i,sep=":"))
} #loop through numbers 1-5
```

```{r}
for (i in Xvec){
  print(i)
} # loops the lements of a vector directly
```

```{r}
for (i in 1:length(Xvec)){
  print(Xvec[i])
} # indicates the elements of a vector
```

### In each case there is a vector and the loop goes through each cell in the vector. The i variable is an object that gets eplaced with a new number in each loop iteration.


## It may be useful to include a counter variable

```{r}
count1<-1
count10<-1

for(i in 1:5){
  print(paste("count1 =", count1))
  print(paste("count10 =", count10))
  count1 <- count1+1
  count10 <- count10*10
}

```

```{r}
countbefore <- 0
countafter <- 0

for(i in 1:5){
  countbefore <- countbefore + 1
  print(paste("before =", countbefore))
  print(paste("after=",countafter))
  countafter <- countafter + 1
}
```

## Nested loops are loops inside of another, and must have counter variable placed in specific spots.


```{r}
LoopCount <- 0

for(i in 1:3){
  for(j in 1:2){
    LoopCount <- LoopCount+1
    print(paste("Loop =", LoopCount))
    print(paste("i = ",i))
    print(paste("j = ",j))
  }
}
```

## The while(){} loop is interated until  some condition is met inside of a loop - the exit condition

## In bio, this is often used in optimization simulations untiol some optimum/threshold is reached (ex. equilibriums (Evolutionarily Stable Strategy), pop growth trajectories, mutation-selection equilibrium, etc.)

## However, a exit condition may never be reached, so the computer runs an infiinite number of loops.

```{r}
count <- 0
while(count < X){
  print(count)
  count <- count+1
}
```

```{r}
## %% returns the remainder of a division equation.

for(i in 1:9){
  if(i %% 3 ==0 ){
    print(paste("Iteration:",i))
  }
}# divide by N and determine if the dividend is 0
```


# Some loops are very slow to run, so eventually faster version should be found.


```{r}
## A slow loop demonstrating the Central Limit Theorem (sampling 1000 number from a random normal distribuition and getting the average, repeat 5000X, and get the average of all these)


Iters <- 500 ### Number of iterations
OutVector <- NA
Start <- Sys.time()
for(i in 1:Iters){
  TempMean <- NA
  for(j in 1:1000){ ### One loop per sample
    TempMean[j] <- rnorm(1)
  }
  OutVector[i] <- mean(TempMean)
}
Sys.time()-Start

paste("Mean of means =",mean(OutVector))
```

```{r}
Iters <- 500
OutVector <- vector("numeric", Iters)
Start <- Sys.time()
for(i in 1:Iters){
  TempMean <- vector("numeric",1000)
  for(j in 1:100){ #One loop per sample
    TempMean[j] <- rnorm(1)
  }
  OutVector[i] <- mean(TempMean)
}
Sys.time() - Start

paste("Mean of means =", mean(OutVector))
```


```{r}
## Even faster loop, using sapply() or tapply()

Iters <- 500
OutVector <- vector("numeric", Iters)
OutMean <- function(x){
  return(mean(rnorm(1000)))
}
Start <- Sys.time()
OutVector <- sapply(OutVector, FUN = OutMean)
Sys.time() - Start

paste(mean(OutVector))
```
















